Estimating duration using survival analysis

### 2023-10-02 cis-perfect-testing.tex
Finish chapter rewrite/edit

The estimation of incidence from prevalence is sensitive to the tail of the duration distribution.
In \cref{E-ATACCC} I estimated the duration distribution from the ATACCC dataset, however, this study suffered from a small sample size and only 20 days of follow-up.
Therefore, the estimation of the tail of the duration distribution is driven by model assumptions.
This chapter develops methods to estimate the distribution using CIS data, which has less frequent sampling but a large sample size and long follow-up.

I take a Bayesian survival analysis approach to model the coarse sampling of the CIS (see \cref{perf-test:sec:problem}).
Flat priors can be unintentionally very informative, and hence uninformative priors should be used.
Of particular interest is a strongly informative prior to combine the information in the CIS with the results of \cref{E-ATACCC} (see \cref{perf-test:sec:parameters-priors}).
Through a simulation study, I show that these methods perform well, and in this context are not sensitive to model or whether the prior is vague or strongly informative (see \cref{perf-test:sec:simulation-study}).
However, application in reality must account for false negatives, the subject of \cref{E-imperf-test}.

@@

The CIS infrequently tests each individual, with tests occurring either weekly or monthly (see\todo{refer to section with CIS intro}).
The infrequent testing complicates analysis of the data in two ways: infection episodes can go undetected and the infection episode endpoints (start and end times) are known imprecisely.
This section explains these issues fully, and the remainder of the chapter develops a method to model them.

The first complication in the analysis is that infection episodes can be undetected, meaning no there are no positive tests associated with them.
This happens when an episode starts and ends between tests (see \cref{perf-test:fig:truncation}).
For these undetected episodes, there is no information on them episode, meaning that we do not even know how many of them occurred.
It is important to account for undetected episodes because they are systematically of shorter duration than detected episodes.
To see why this is the case, consider the undetected infection in the bottom of \cref{perf-test:fig:truncation}.
The infection episode depicted begins at time 33 and has a 10-day duration.
If the duration was 24 days (or longer), the individual would still have been detectable at their next test, and therefore would have been detected.
Undetected infections are a large problem in CIS because the interval between tests is often 28 days (see\todo{ref section on the survey design}), large relative to the duration of an infection episode.
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics[width=1.2\textwidth]{cis-perfect-testing/truncation}}
  \caption[Undetected episodes in CIS data]{An unknown number of infection episodes in CIS are undetected. Consider the two individuals shown here: we collect identical data for them (a series of negative tests) yet the top individual was never infected and the bottom individual was. See the main text for why this is an issue. \label{perf-test:fig:truncation}}
\end{figure}

The second complication is that the beginning and end times of detected infection episodes are known imprecisely.
The imprecision in the beginning of the infection episode arises because the only observation that an episode has started is that an individual was negative at one test, on day $l_i^{(b)}-1$ (the minus one is because the first day that the infection could have begun is the day after the negative test), and positive at the next test, on day $r_i^{(b)}$.
Similarly, the end of the infection episode is bounded by the final positive test on day $l_i^{(e)}$ and the following negative test on day $r_i^{(e)}+1$ (the plus one is because the last day that the infection could have ended is the day before the negative test).
\Cref{perf-test:fig:double-interval-censor} shows this issue graphically.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/double-interval-censor}
  \caption[Double-interval censoring in CIS data]{Episodes data in CIS is double interval censored, meaning that both the start and end of the episode are only known up to an interval. Demonstrated here by a participant who is recorded negative at time 7 to positive at time 14, bounding the start of the episode; similarly, changing from positive at time 28 to negative at time 56 bounds the end of the episode. \label{perf-test:fig:double-interval-censor}}
\end{figure}

I will base my analysis on the 4,800\todo{check number} of episodes that were first detected between 10th Oct 2020 and 6th Dec 2020 inclusive and have negatives bounding the start and end time of the episode.
That is, the individual in which the infection episode occurred tested negative at least once in the survey prior to the start of their infection episode, and they tested negative at least once after their infection episode ended.
Without a negative before the episode there is little information on the episode's length because it's start time could have occurred arbitrarily long ago.
The only episodes without a negative following the end of the episode are those in individuals lost to follow-up before their episode ended, which is extremely rare.
This filtering could bias the results of the study but will be considered in the same way as undetected episodes to avoid this.

In the remainder of this and the following chapter, I use survival analysis rather than the biomarker approach taken in \cref{E-ATACCC} for computational efficiency and data limitations arising from the study design.
In the biomarker approach, computational efficiency is an issue because the number of random effects scales with the number of detected infection episodes.
There are a large number of detected episodes in CIS; in the period I will consider, around 50 times more than in ATACCC.
Therefore, the biomarker approach would be computationally challenging with the large sample size in CIS.
The computational challenge is compounded by the data being stored in a TRE with the associated limited access to computational power and software.
The study design of CIS, with the infrequent sampling in the CIS study design, means that most detected episodes have only a single positive test.
Hence, there is no information on the viral load trajectory required to identify the random effects in the model used in \cref{E-ATACCC} or similar approaches.
While the biomarker approach does use more information per sample by taking into account the viral load as well as the binary test result (\ie whether the result is positive or negative), however, the available information is unlikely to be sufficient to overcome these problems.

@@

In survival analysis, the event at the start of the time of interest is the \emph{initiating event}, and the event at the end of the time of interest is the \emph{terminating event}.
It is broadly applicable to many domains, such as the time-to-failure for a mechanical system or the effect of a treatment on time spent in hospital.
In the context of this thesis, the distribution to estimate is the duration of positivity.
Therefore, the initiating event is when the duration starts; and the terminating event is when the duration ends.

Undetected events where we have information on missed events is unusual within the survival analysis literature.
Far more common is that there is no data on these events, a situation known as \emph{truncation}.
A situation that is particularly relevant is \emph{left truncation}, when only events lasting a minimum amount of time are observed.
A subtlety that makes the CIS different from standard left truncation is that information on individuals in which undetected infections occurred is available because they were enrolled in the study.
In standard left truncation, nothing is known about the missed events.
As the information is not being incorporated here, standard left truncation methods remain applicable, and they can be modified to incorporate the additional information if desired.

The time of the initiating and/or terminating events are not known exactly.
This is known as \emph{censord} in the survival analysis literature.
Censored events are present in almost all survival analyses.
An event is \emph{left censored} if it is known to occur before a certain time, \emph{right censored} if it is known to occur after a certain time, and \emph{interval censored} if it is known to occur within an interval.
%Left and right censoring can be viewed as special cases of interval censoring where the lower bound is negative infinity or the upper bound is positive infinity respectively.
Interval censoring arises in CIS because we only observe the change from being not detectable (\ie: negative) at one test, to detectable (\ie: positive) at the following test and vice versa (see \cref{perf-test:fig:double-interval-censor}).
If both the initiating and terminating events are censored, then the event is \emph{doubly censored}.
Doubly censored data is more challenging to handle because the initiating event time and duration must be jointly modelled~\autocite[and references therein]{liSemiparametric}.
In CIS data both the initiating and the terminating events are interval censored.
Therefore, the CIS data is \emph{double interval censored}; approaches for double interval censoring are reviewed by \textcite{sunAnalysis,bogaertsSurvival}.

Few studies consider the combination of both double interval censoring and undetected or truncated events
This is especially true within the human biostatistical literature.
While this literature does contain theoretical frameworks which include the double censored and truncated case~\autocites{turnbullEmpirical}{dempsterMaximum}, they have been applied when the terminating event was either uncensored or right censored~\autocite{sunEmpirical,bacchettiNonparametric}.
The approach of these studies \autocite[and elsewhere, e.g.:][]{shenNonparametric} was to take a conditional likelihood approach, conditioning on both the truncation (a selection effect) and the period in which the initiating event is known to occur.
They justify conditioning on the time of the initiating event by arguing that this information provides little information on the distribution being estimated~\citePersonalComms{Nick Jewell}.
However, this is not the case here.
In CIS, infection episodes are more likely to be detected when an individual is in the weekly testing phase.
This phase is not representative of the overall study: infections are more likely to be detected during this phase.
Therefore, estimating the probability of being detected conditional on being infected in this period (which would be implictly done in this previous method)  will overestimate the probability of being detected.

Censored and truncated data frequently occur when estimating the survival time of bird nests, with the censoring taking a variety of forms~\autocite{heiseyABCs}.
In this context, \textcite{heiseyModelling} develop a general framework allowing for arbitrary censoring patterns and double interval censoring.
In their application, the visit times are global (all nests could be detected on the same days), while in CIS each individual has their own visit schedule; with global visit times the probability of being detected is the same for all participants in the study, simplifying the likelihood.
I use this framework as the basis of the model I derive in \cref{perf-test:sec:model}.

Inference will be performed within the Bayesian paradigm for two reasons.
First, the standard frequentist methods for estimating the parameter's variance and covariance are poorly developed in this area.
Commonly used methods are based on the Fisher information matrix or resampling, but neither have theoretical justification~\autocite[230]{sunAnalysis}\todo{Try and understand why this is the case}.
Second, a prior naturally allows the inclusion information from previous studies, such as that from \cref{E-ATACCC}, and/or a belief that the hazard changes smoothly in a way that can be encoded in a prior (these options are explored in \cref{perf-test:sec:parameters-priors}).
These prior structures may be needed to combat the lack of information in the data due to the undetected episodes~\autocite{caoBias}.
They can also be viewed as combining the information from the two analyses (see \cref{perf-test:sec:informative-priors}).

Bayesian methods have previously been applied to nest survival settings~\autocites{heBayesiana}{heBayesian}{caoModeling}, augmenting the data with the unobserved times of infection.
The augmentation creates a parameter per detected infection.
Therefore, data augmentation applied to the 4000 detected episodes in the main analysis of \cref{E-imperf-test} greatly increases the dimensionality of the problem, making it computationally infeasible.
I develop a Bayesian framework where there is no explicit data augmentation, allowing the method to scale to the large number of episodes present in CIS.

@@

\subsection{Likelihood}\label{perf-test:sec:likelihood}

@@

Integrating over the number of missed episodes is analytically tractable making inference tractable within modern, high-performance Bayesian software such as Stan.
Let the total number of infection episodes in the cohort be $\ntot = \ndet + \nnodet$ where $\ndet$ is the number of detected infections and $\nnodet$ is the number of undetected infections.
There are $\Ncis$ individuals in the cohort.
Without loss of generality, let the individuals with a detected episode be indexed by $i = 1, \dots, \ndet$ and the individuals without a detected episode be indexed by $i = \ndet + 1, \dots, \Ncis$.
Over the short period of time considered, no individual has more than one detected episode. 
Let $j = 1, \dots \ntot$ index infection episodes, ordered such that for $j \leq \ndet$ the $j$th infection episode occurs in the $i$th individual.

All the relevant information about episode $j$ can be fully characterised by the (unobserved) triplet $(b_j, e_j, \sched_j)$ where $b_j$ and $e_j$ are the times the episode being and end respectively and $\sched_j$ is the schedule of tests for the individual in which the episode occurs. 
If $\sched_j$ contains at least one test between $b_i$ and $e_i$ (inclusive), then the episode is detected, otherwise it is undetected.
The triplet belongs to a state space where $b_j$ and $e_j$ span possible infection and recovery times, and $\sched_j$ is one of the observed test schedules (assuming other test schedules occur with probability zero).

To relate this to the data, define three sets which partition the space of possible infection and recovery times (\ie: are strict subsets of $T \times T$), for individual $i$ (graphically shown in Figure \ref{perf-test:fig:partitionSpace}).
Each episode, whether detected or not, must fall into one of these three classes with a probability that is a function of the model parameters, $\theta$.

@@

Any episode $j$ independently belongs to one of these classes.
I assume the individual in which each episode occurs is iid uniform across the cohort.
Therefore, conditional on $\ntot$, the number of episodes in each class (that is, the counts $n_{ia}, n_{iu}, n_{i\bar{a}}$) are distributed multinomially.
The above probabilities ($p_{iu}$, $p_{ia}$, and $p_{i\bar{a}}$) are conditional on the episode occurring in individual $i$.
These probabilities are multiplied by $1/\Ncis$ to give the probability that an episode belongs to each class, without conditioning on a specific individual.

@@

\todo{better phrasing needed for this paragraph, mauybe try email to Shaun?}
In forming the probability $p_{ia}$, the data is used.
This is not quite correct because forming a likelihood should not depend on the data.
However, an alternative where each combination of infection and recovery times is considered its own class is possible.
Then, the counts for all the classes that previously contained undetected episodes would be zero.
The resulting likelihood would be the same.
Therefore, I present this version as it is clearer.

The posterior we are interested in is $p(\theta \mid \na)$ where $\theta$ are the parameters governing the survival distribution.
The probabilities $\pcis$ are functions of the parameters $\theta$.
\begin{align}
&p(\theta \mid \na) \\
&\propto p(\theta) p(\na \mid \theta) \\
&= p(\theta) \sum_{\ntot=\ndet}^\infty p(\ntot \mid \theta) p(\na \mid \theta, \ntot) \\
&\propto p(\theta) \sum_{\ntot=\ndet}^\infty \left( p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{i=\ndet}^{\ndet} p_{ia} \right) \\
&= p(\theta) \left( \prod_{i=\ndet}^{\ndet} p_{ia} \right) \left( p(\ntot \mid \theta) \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \right) \label{perf-test:eq:full-posterior}.
\end{align}

In the remainder of the section, expressions for each $p_{ia}$, $p_{iu}$, and analytical solutions for the sum in \cref{perf-test:eq:full-posterior} are derived in terms of the observed data, under different priors.

I derive an analytical solution to $p(\ntot \mid \theta) = \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}$ under a negative binomial prior for $\ntot$.
This prior is $\NegBin(\mu, r)$ using the mean/dispersion parameterisation of the negative binomial.
Equivalently:
\begin{align}
\ntot \mid \lambda &\dist \Poi(\lambda) \\
\lambda &\dist \GamDist(a, b)
\end{align}
where $b = r / \mu$ and $a = r$.
Hence:
\begin{align}
&\sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} p(\ntot) \\
&= \int \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} p(\ntot \mid \lambda) p(\lambda) d\lambda &\text{$\lambda$ explicit}\\
&= \int \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} \frac{\lambda^{\ntot} e^{-\lambda}}{\ntot!} p(\lambda) d\lambda &\ntot \dist \Poi\\
%&= \int \sum_{\ntot=\ndet}^\infty \frac{1}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} \lambda^{\ntot-\ndet} \lambda^{\ndet} e^{-\lambda} p(\lambda) d\lambda \\
&= \int \lambda^{\ndet} e^{-\lambda} p(\lambda) \sum_{\nnodet=0}^\infty \frac{(\pnodet \lambda)^{\nnodet}}{\nnodet!} d\lambda &\nnodet = \ntot-\ndet\\
&= \int \lambda^{\ndet} e^{-\lambda} p(\lambda) e^{\lambda \pnodet} d\lambda &\text{Maclaurin series of $e$} \\
&= \int \lambda^{\ndet} e^{-\lambda(1 - \pnodet)} \frac{b^a}{\Gamma(a)} \lambda^{a-1} e^{-b\lambda} \lambda d\lambda &\text{Gamma pdf}\\
&= \int \frac{b^a}{\Gamma(a)} \lambda^{a+\ndet-1} e^{-(b+1-\pnodet)\lambda} \lambda d\lambda \\
&= \frac{b^a}{\Gamma(a)} \frac{\Gamma(a+\ndet)}{(b+1-\pnodet)^{a+\ndet}} &\text{Gamma pdf}\\
&\propto (b+1-\pnodet)^{-(a+\ndet)} \\
&= (r/\mu + 1 - \pnodet)^{-(r+\ndet)} &\text{sub in $\mu$ and $r$}\\
&\propto(r + \mu (1- \pnodet))^{-(r+\ndet)}.
\end{align}

Substituting this into \cref{perf-test:eq:full-posterior} gives the following expression for the posterior.
\begin{align}
p(\theta \mid \na)
&\propto p(\theta) \left( \prod_{i=\ndet}^{\ndet} p_{ia} \right) (r + \mu (1- \pnodet))^{-(r+\ndet)} \label{perf-test:eq:full-posterior-simplified}.
\end{align}

$\ntot$ is generally a nuisance parameter, however, its posterior distribution can be useful for diagnostic purposes (\eg: see X\todo{ref where I use this in the next chapter}).
I reconstruct the posterior using a posterior sample of $\theta$.
For each posterior sample of $\theta$, I sample from the full conditional of $\ntot$.
This gives the joint posterior of $\ntot$ and $\theta$.
The full conditional of $\ntot$ is given by:
\begin{align}
p(n_\text{tot} \mid \ncis, \theta)
&\propto p(\ntot \mid \theta) p(\ncis \mid \theta, \ntot) \\
&\propto p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} &\text{by \cref{perf-test:eq:augmented-likelihood}} \\
&\propto \frac{\Gamma(r + \ntot)}{\ntot!} \left( \frac{\mu}{r + \mu} \right)^{\ntot} \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot} \\
&= \frac{\Gamma(r + \ntot)}{(\ntot-\ndet)!} \left( \frac{\mu \pnodet}{r + \mu} \right)^{\ntot}  \\
&= \frac{\Gamma((r + \ndet) + (\ntot- \ndet))}{(\ntot-\ndet)!} \left( \frac{\mu \pnodet}{r + \mu} \right)^{\ntot-\ndet}.
\end{align}
Comparing this expression to the pmf of a negative binomial, we find that $\ntot - \ndet$ is distributed negative binomial with size parameter $r+\ndet$ and probability parameter $(r + \mu(1 - \pnodet))/(r + \mu)$.
The mean of this distribution is $(r+\ndet)\mu \pnodet/r+\mu(1-\pnodet)$.

>>>>

The estimation of incidence from prevalence is sensitive to the tail of the duration distribution.
In \cref{E-ATACCC} I estimated the duration distribution from the ATACCC dataset.
However, this study suffered from a small sample size and only 20 days of follow-up.
Therefore, model assumptions determined the size of the tail estimated.
In this chapter, I develop a method to estimate the distribution using CIS data. 
CIS sampling is less frequent, but with a large sample size and long follow-up.

I take a Bayesian survival analysis approach to model the coarse sampling of the CIS (see \cref{perf-test:sec:problem}).
Flat priors are unintentionally very informative.
Hence, at least weakly informative priors should be used.
Of particular interest is a strongly informative prior to combine the information in the CIS with the results of \cref{E-ATACCC} (see \cref{perf-test:sec:parameters-priors}).
I conduct a simulation study to demonstrate this method's performance.
It succesfully recovers the truth when using reasonable priors (see \cref{perf-test:sec:simulation-study}).
In reality, the method must account for false negatives, discussed in \cref{E-imperf-test}.

@@

The CIS tests each individual infrequently, either weekly or monthly (see\todo{refer to section with CIS intro}).
The infrequent testing complicates analysis of the data in two ways.
First, infection episodes can go undetected.
Second, the infection episode start and end times are known imprecisely.
This section explains these issues.
The rest of the chapter develops a method to model them.

The first complication in the analysis is that infection episodes can be undetected.
This means there are no positive tests associated with them.
Undetected episodes occur when an episode starts and ends between tests (see \cref{perf-test:fig:truncation}).
For undetected episodes, there is no information on them episode; even the number that occurred is unknown.
Undetected episodes are important because they are shorter than detected episodes.
To see why this is the case, consider the undetected infection in the bottom of \cref{perf-test:fig:truncation}.
The infection episode depicted begins at time 33 and has a 10-day duration.
If the duration was 24 days or more, the individual would still be detectable at their next test and detected then.
CIS has many undetected infections. This is because frequency of testing (up to 28 days, see\todo{ref section on the survey design}) is longer than most infections (median around 15 days).
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics[width=1.2\textwidth]{cis-perfect-testing/truncation}}
  \caption[Undetected episodes in CIS data]{An unknown number of infection episodes in CIS are undetected. Consider the two individuals shown here: we collect identical data for them (a series of negative tests) yet the top individual was never infected and the bottom individual was. See the main text for why this is an issue. \label{perf-test:fig:truncation}}
\end{figure}

The second complication is that the exact beginning and end times of detected infection episodes are unknown.
The imprecision in the beginning of the infection episode arises because the only observation that an episode has started is that an individual was negative at one test, on day $l_i^{(b)}-1$ (the minus one is because the first day that the infection could have begun is the day after the negative test), and positive at the next test, on day $r_i^{(b)}$.
Similarly, the end of the infection episode is bounded by the final positive test on day $l_i^{(e)}$ and the following negative test on day $r_i^{(e)}+1$ (the plus one is because the last day that the infection could have ended is the day before the negative test).
\Cref{perf-test:fig:double-interval-censor} shows this issue graphically.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/double-interval-censor}
  \caption[Double-interval censoring in CIS data]{Episodes data in CIS is double interval censored, meaning that both the start and end of the episode are only known up to an interval. Demonstrated here by a participant who is recorded negative at time 7 to positive at time 14, bounding the start of the episode; similarly, changing from positive at time 28 to negative at time 56 bounds the end of the episode. \label{perf-test:fig:double-interval-censor}}
\end{figure}

I base my analysis on the 4,800\todo{check number} episodes that were first detected between 10th Oct 2020 and 6th Dec 2020 inclusive with negatives bounding the start and end time of the episode.
That is, the individual tested negative before and after the episode.
Without a negative before the episode there is no lower bound on the episode's start time.
It could be arbitrarily long and there is little information on its length.
A negative test follows each episode unless the individual was lost to follow-up before their episode ended.
This is rare.
Episodes removed in this step are incorporated in the adjustment for undetected episodes.

In the rest of this and the following chapter, I use survival analysis rather than the biomarker approach taken in \cref{E-ATACCC}.
This choice is motivated by computational efficiency and limitations of the CIS's study design.
The biomarker approach is computational challenging because the number of random effects scales with the number of detected infection episodes.
There are many detected episodes in CIS; in the period I will consider, around 50 times more than in ATACCC.
Therefore, the biomarker approach would be computationally challenging.
The computational challenge is compounded by the data being stored in a TRE, which has limited computational power and software.
The issue with random effects is made worse by the CIS's study design.
Infrequent testing means that most detected episodes have only a single positive test.
Hence, there is no information on the viral load trajectory and the random effects are unidentifiable.

@@

In survival analysis, the event at the start of the time of interest is the \emph{initiating event}, and the event at the end of the time of interest is the \emph{terminating event}.
It is used in many areas, including how long a machine lasts before breaking down or how long a patient spends in hospital after a treatment.
Here, I am estimating the distribution of the duration of positivity.
The initiating event is the episode's start; the terminating event is the episode's end.

Undetected events where there is no information on these events is known as \emph{truncation}.
A situation that is particularly relevant is \emph{left truncation}, when only events lasting a minimum amount of time are observed.
There is a subtlety that makes the CIS different from standard left truncation.
In CIS, individuals in which undetected infections occurred are enrolled in the study.
Therefore, some data has been collected about them.
In standard left truncation, nothing is known about the missed events.
I incorporate the additional information into the analysis.
I refer to these events as ``undetected'' to emphasise this difference.

The time of the initiating and/or terminating events are not known exactly.
This is known as \emph{censored} in the survival analysis literature.
Censored events are present in almost all survival analyses.
An event is \emph{left censored} if it is known to occur before a certain time.
An event is \emph{right censored} if it is known to occur after a certain time.
An event is \emph{interval censored} if it is known to occur within an interval.
%Left and right censoring can be viewed as special cases of interval censoring where the lower bound is negative infinity or the upper bound is positive infinity respectively.
As explained in \cref{perf-test:sec:problem}, the initiating and terminating events are only bounded, and hence interval censored.
The data is \emph{doubly censored} because both the initiating and terminating events are censored.
To handle doubly censored data, the times need to be jointly modelled~\autocite[and references therein]{liSemiparametric}.
In CIS data both the initiating and the terminating events are interval censored.
Therefore, the CIS data is \emph{double interval censored}.
\Textcite{sunAnalysis,bogaertsSurvival} review methods for double interval censored data.

Few studies look at the combination of both double interval censoring and undetected (or truncated) events
This is especially true within the human biostatistical literature.
The literature does include theoretical frameworks which include the double censored and truncated case~\autocites{turnbullEmpirical}{dempsterMaximum}.
The frameworks have only been applied when the terminating event was either uncensored or right censored~\autocite{sunEmpirical,bacchettiNonparametric}.
These studies \autocite[and elsewhere, e.g.:][]{shenNonparametric} take a conditional likelihood approach.
They condition on the truncation (a selection effect) and the period where the initiating event occurs.
They condition on the initiating event time because it provides negligible information on the distribution of interest~\citePersonalComms{Nick Jewell}.
This is not the case for the CIS.
In CIS, infection episodes are more likely to be detected when an individual is in the weekly testing phase.
CIS is more likely to detect infection episodes during the weekly testing phase.
Therefore, the probability of an infection episode being detected conditional on being infected is higher than the overall probability of being detected.\todo{this is a bit unclear, try and rephrase?}

Censored and truncated data are normal for estimation of the survival time of bird and insect nests~\autocite{heiseyABCs}.
\textcite{heiseyModelling} developed a framework for handling this situation.
The framework is general, allowing for arbitrary censoring patterns and double interval censoring.
They apply it to a situation the visit times are global (all nests could be detected on the same days), simplifying the likelihood.
I adapt this framework for CIS, when each individual has their own visit schedule (see section \cref{perf-test:sec:model}).

I use Bayesian inference for two reasons.
First, frequentist methods for estimating the parameter's variance and covariance are poorly developed.
Commonly used methods are based on the Fisher information matrix or resampling.
Neither have theoretical justification~\autocite[230]{sunAnalysis}\todo{Try and understand why this is the case}.
Second, a prior can include information from previous studies, such as from \cref{E-ATACCC}.
Alternatively, the prior can favour a smoothly changing hazard.
I explore priors in section 1.4.
These prior structures overcome issues caused by the undetected episodes ~\autocite{caoBias}.

Bayesian methods been used for nest survival before~\autocites{heBayesiana}{heBayesian}{caoModeling}, augmenting the data with the unobserved times of infection.
The augmentation creates a parameter per detected infection.
This increases the dimensionality of the problem greatly, making it computationally infeasible here.
I develop a novel Bayesian framework without explicit data augmentation.
This method scales to the large number of episodes present in CIS.

@@

@@

Integrating over the number of missed episodes is analytically tractable.
This makes inference tractable within modern, high-performance Bayesian software such as Stan.
Let the total number of infection episodes in the cohort be $\ntot = \ndet + \nnodet$ where $\ndet$ is the number of detected infections and $\nnodet$ is the number of undetected infections.
Let $\Ncis$ be the number of individuals in the cohort.
Without loss of generality, let the individuals with a detected episode be indexed by $i = 1, \dots, \ndet$ and the individuals without a detected episode be indexed by $i = \ndet + 1, \dots, \Ncis$.
Over the short period of time considered, no individual has more than one detected episode. 
Let $j = 1, \dots \ntot$ index infection episodes, ordered such that for $j \leq \ndet$ the $j$th infection episode occurs in the $i$th individual.

All the relevant information about episode $j$ can be fully characterised by the (unobserved) triplet $(b_j, e_j, \sched_j)$ where $b_j$ and $e_j$ are the times the episode being and end respectively and $\sched_j$ is the set of test times for the individual in which the episode occurs. 
If $\sched_j$ contains at least one time between $b_i$ and $e_i$ (inclusive), one time prior to $b_i$, and one time after $e_i$, then the episode is detected, otherwise it is undetected.
The triplet belongs to a state space.
$b_j$ and $e_j$ span possible infection and recovery times.
$\sched_j$ is one of the observed test schedules, with any other test schedule assumed to occur with probability zero.

To relate the state space to the data, I define three subsets partitioning the space of possible infection and recovery times (\ie: are strict subsets of $T \times T$), for any individual based on their test times.
Figure \ref{perf-test:fig:partitionSpace} shows this graphically.
Each episode, whether detected or not, must fall into one of these three classes with a probability that is a function of the model parameters, $\theta$.

@@

Any episode $j$ independently belongs to one of these classes.
I assume the individual in which each episode occurs is iid uniform across the cohort.
Therefore, conditional on $\ntot$, the number of episodes in each class, $(n_{ia}, n_{iu}, n_{i\bar{a}})^T$, are distributed multinomially.
The above probabilities ($p_{iu}$, $p_{ia}$, and $p_{i\bar{a}}$) are conditional on the episode occurring in individual $i$.
These probabilities are multiplied by $1/\Ncis$ to give the probability that an episode belongs to each class, without conditioning on a specific individual.

@@

\todo{Not sure if the followin paragraph is understandable}
In forming the probability $p_{ia}$, the data is used.
This is not quite correct because forming a likelihood should not depend on the data.
However, an alternative where each combination of infection and recovery times is considered its own class is possible.
Then, the counts for all the classes that previously contained undetected episodes would be zero.
The resulting likelihood would be the same.
Therefore, I present this version as it is clearer.

The posterior we are interested in is $p(\theta \mid \na)$.
The posterior density can be written as follows.
\begin{align}
&p(\theta \mid \na) \\
&\propto p(\theta) p(\na \mid \theta) \\
&= p(\theta) \sum_{\ntot=\ndet}^\infty p(\ntot \mid \theta) p(\na \mid \theta, \ntot) \\
&\propto p(\theta) \sum_{\ntot=\ndet}^\infty \left( p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{i=\ndet}^{\ndet} p_{ia} \right) \\
&= p(\theta) \left( \prod_{i=\ndet}^{\ndet} p_{ia} \right) \left( p(\ntot \mid \theta) \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \right) \label{perf-test:eq:full-posterior}.
\end{align}
The rest of this section derives expressions for each $p_{ia}$, $p_{u}$ and the sum.

Now I derive an analytical solution to $p(\ntot \mid \theta) = \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}$ under the prior $\ntot \dist \NegBin(\mu, r)$ (using the mean/dispersion parameterisation of the negative binomial).
Equivalently:
\begin{align}
\ntot \mid \lambda &\dist \Poi(\lambda) \\
\lambda &\dist \GamDist(a, b)
\end{align}
where $b = r / \mu$ and $a = r$.
Hence:
\begin{align}
&\sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} p(\ntot) \\
&= \int \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} p(\ntot \mid \lambda) p(\lambda) d\lambda &\text{$\lambda$ explicit}\\
&= \int \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} \frac{\lambda^{\ntot} e^{-\lambda}}{\ntot!} p(\lambda) d\lambda &\ntot \dist \Poi\\
%&= \int \sum_{\ntot=\ndet}^\infty \frac{1}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} \lambda^{\ntot-\ndet} \lambda^{\ndet} e^{-\lambda} p(\lambda) d\lambda \\
&= \int \lambda^{\ndet} e^{-\lambda} p(\lambda) \sum_{\nnodet=0}^\infty \frac{(\pnodet \lambda)^{\nnodet}}{\nnodet!} d\lambda &\nnodet = \ntot-\ndet\\
&= \int \lambda^{\ndet} e^{-\lambda} p(\lambda) e^{\lambda \pnodet} d\lambda &\text{Maclaurin series of $e$} \\
&= \int \lambda^{\ndet} e^{-\lambda(1 - \pnodet)} \frac{b^a}{\Gamma(a)} \lambda^{a-1} e^{-b\lambda} \lambda d\lambda &\text{Gamma pdf}\\
&= \int \frac{b^a}{\Gamma(a)} \lambda^{a+\ndet-1} e^{-(b+1-\pnodet)\lambda} \lambda d\lambda \\
&= \frac{b^a}{\Gamma(a)} \frac{\Gamma(a+\ndet)}{(b+1-\pnodet)^{a+\ndet}} &\text{Gamma pdf}\\
&\propto (b+1-\pnodet)^{-(a+\ndet)} \\
&= (r/\mu + 1 - \pnodet)^{-(r+\ndet)} &\text{sub in $\mu$ and $r$}\\
&\propto(r + \mu (1- \pnodet))^{-(r+\ndet)}.
\end{align}

Substituting this into \cref{perf-test:eq:full-posterior} gives:
\begin{align}
p(\theta \mid \na)
&\propto p(\theta) \left( \prod_{i=\ndet}^{\ndet} p_{ia} \right) (r + \mu (1- \pnodet))^{-(r+\ndet)} \label{perf-test:eq:full-posterior-simplified}.
\end{align}

$\ntot$ is generally a nuisance parameter, however, its posterior distribution can be useful for diagnostic purposes (as in X\todo{ref where I use this in the next chapter}).
Its posterior can be reconstructed using a posterior sample of $\theta$.
For each posterior sample of $\theta$, sample from the full conditional of $\ntot$ to give the joint posterior of $\ntot$ and $\theta$.
The full conditional of $\ntot$ is given by:
\begin{align}
p(n_\text{tot} \mid \ncis, \theta)
&\propto p(\ntot \mid \theta) p(\ncis \mid \theta, \ntot) \\
&\propto p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} &\text{by \cref{perf-test:eq:augmented-likelihood}} \\
&\propto \frac{\Gamma(r + \ntot)}{\ntot!} \left( \frac{\mu}{r + \mu} \right)^{\ntot} \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot} \\
&= \frac{\Gamma(r + \ntot)}{(\ntot-\ndet)!} \left( \frac{\mu \pnodet}{r + \mu} \right)^{\ntot}  \\
&= \frac{\Gamma((r + \ndet) + (\ntot- \ndet))}{(\ntot-\ndet)!} \left( \frac{\mu \pnodet}{r + \mu} \right)^{\ntot-\ndet}.
\end{align}
Comparing this expression to the pmf of a negative binomial, we find that $\nnodet = \ntot - \ndet$ is distributed negative binomial with size parameter $r+\ndet$ and probability parameter $(r + \mu(1 - \pnodet))/(r + \mu)$.
%The mean of this distribution is $(r+\ndet)\mu \pnodet/(r+\mu(1-\pnodet))$.
